---
title: '5\. Worksheet: Alpha Diversity'
author: "Elaine Hoffman; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin = 2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of local or site-specific diversity, also known as alpha ($\alpha$) diversity. 
First we will quantify two of the fundamental components of ($\alpha$) diversity: **richness** and **evenness**.
From there, we will then discuss ways to integrate richness and evenness, which will include univariate metrics of diversity along with an investigation of the **species abundance distribution (SAD)**.

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) to your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Answer questions in the worksheet.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme).
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For the assignment portion of the worksheet, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file `AlphaDiversity_Worskheet.Rmd` and the PDF output of `Knitr` (`AlphaDiversity_Worskheet.pdf`).

## 1) R SETUP

In the R code chunk below, please provide the code to: 
1) Clear your R environment,
2) Print your current working directory,
3) Set your working directory to your `Week-2/` folder folder, and
4) Load the `vegan` R package (be sure to install first if you have not already).

```{r}
rm(list = ls())
getwd()
setwd("/cloud/project/Week2-Alpha")

require("vegan")

```

## 2) LOADING DATA

In the R code chunk below, do the following:
1) Load the BCI dataset, and 
2) Display the structure of the dataset (if the structure is long, use the `max.level = 0` argument to show the basic information).

```{r}
data("BCI")
help(BCI)
str(BCI)

```

## 3) SPECIES RICHNESS

**Species richness (S)** refers to the number of species in a system or the number of species observed in a sample.

### Observed richness

In the R code chunk below, do the following:

1. Write a function called `S.obs` to calculate observed richness

2. Use your function to determine the number of species in `site1` of the BCI data set, and

3. Compare the output of your function to the output of the `specnumber()` function in `vegan`.

```{r}

S.obs <- function(x = ""){
   rowSums(x > 0) * 1
}

site1 <- BCI[1, ]
S.obs(site1)
specnumber(site1)
site1t4 <- BCI[1:4,]
S.obs(site1t4)

```

***Question 1***: Does `specnumber()` from `vegan` return the same value for observed richness in `site1` as our function `S.obs`?
What is the species richness of the first four sites (i.e., rows) of the BCI matrix?

> ***Answer 1***: Yes, the specnumber() and the S.obs() return the same value for richness in site1. The richness of the first four sites is 93, 84, 90, and 94 respectively. 
>


### Coverage: How well did you sample your site?

In the R code chunk below, do the following:

1. Write a function to calculate Good's Coverage, and

2. Use that function to calculate coverage for all sites in the BCI matrix.

```{r}
C <- function(x = ""){
  1-(rowSums(x == 1) / rowSums(x))
}
C(BCI)

C(site1)


```
  
***Question 2***: Answer the following questions about coverage:

a.  What is the range of values that can be generated by Good's Coverage? 
b.  What would we conclude from Good's Coverage if $n_{i}$ equaled *N*?
c.  What portion of taxa in `site1` was represented by singletons? 
d.  Make some observations about coverage at the BCI plots.

> ***Answer 2a***: Good's coverage yields values between 0 and 1
>

> ***Answer 2b***:If ni equalled N, then every species observation was a singleton which yields a good's coverage value of 0. This indicates that the sample has not been covered thoroughly enough. 
>

> ***Answer 2c***: ~7% of species are represented by singletons.
>

> ***Answer 2d***: Almost all of the coverage values are 90% or higher and none are less than 80%. This supports that there was a thorough coverage of the BCI plots. 
>


### Estimated richness

In the R code chunk below, do the following:

1. Load the microbial dataset (located in the `Week-2/data` folder),

2. Transform and transpose the data as needed (see handout),

3. Create a new vector (`soilbac1`) by indexing the bacterial OTU abundances of any site in the dataset,

4. Calculate the observed richness at that particular site, and 

5. Calculate coverage of that site

```{r}
soilbac <- read.table("data/soilbac.txt", sep = "\t", header = TRUE, row.names = 1)
soilbac.t <- as.data.frame(t(soilbac))
soilbac1 <- soilbac.t[1,]

S.obs(soilbac1)
C(soilbac1)

```

***Question 3***: Answer the following questions about the soil bacterial dataset. 

a.  How many sequences did we recover from the sample `soilbac1`, i.e. *N*? 
b.  What is the observed richness of `soilbac1`? 
c.  How does coverage compare between the BCI sample (`site1`) and the KBS sample (`soilbac1`)?

> ***Answer 3a***:13310
>

> ***Answer 3b***: 1074
>

> ***Answer 3c***: For the BCI sample site1 only had ~7% of observations that were singletons. In comparison, KBS soilbac1 had ~35% of observations that were singletons which indicates that the sampling coverage was not as thorough in the KBS sample. 
>


### Richness estimators

In the R code chunk below, do the following:

1. Write a function to calculate **Chao1**,

2. Write a function to calculate **Chao2**, 

3. Write a function to calculate **ACE**, and

4. Use these functions to estimate richness at `site1` and `soilbac1`. 

```{r}

S.chao1 <- function(x = ""){
  S.obs(x) + (sum(x == 1)^2)/ (2 * sum(x == 2))
}

S.chao2 <- function(site , SbyS){
  SbyS = as.data.frame(SbyS)
  x = SbyS[site, ]
  SbyS.pa <- (SbyS > 0) * 1
  Q1 = sum(colSums(SbyS.pa) == 1)
  Q2 = sum(colSums(SbyS.pa) == 2)
  S.chao2 = S.obs(x) + (Q1^2)/(2*Q2)
  return(S.chao2)
}

S.ace <- function(x = "", thresh = 10){
  x <- x[x>0]
  S.abund <- length(which(x > thresh))
  S.rare <- length(which(x <= thresh))
  singlt <- length(which(x == 1))
  N.rare <- sum(x[which(x <= thresh)])
  C.ace <- 1 - (singlt / N.rare)
  i <- c(1:thresh)
  count <- function(i, y){
    length(y[y == i])
  }
  a.1 <- sapply(i, count, x)
  f.1 <- (i * (i-1)) * a.1
  G.ace <- (S.rare/C.ace)*(sum(f.1)/(N.rare*(N.rare-1)))
  S.ace <- S.abund + (S.rare/C.ace) + (singlt/C.ace) * max(G.ace,0)
  return(S.ace)
}

S.chao1(site1)
S.chao2(1,BCI)
S.ace(site1)
S.chao1(soilbac1)
S.chao2(1,soilbac.t)
S.ace(soilbac1)

```

***Question 4***: 
What is the difference between ACE and the Chao estimators? 
Do the estimators give consistent results? 
Which one would you choose to use and why?

> ***Answer 4***:The Chao estimators use instances of singletons and doubletons in a count to make inferences whereas the ACE estimator uses a threshold of abundance for individuals of a species to be considered rare. For example, ACE often considers species with 10 or less sampled individuals to be rare. This means the ACE threshold for "rareness" is much higher than the Chao estimators. 
> The estimators do not give consistant results. The Chao1 value for site1 is ~119 and the ACE value is ~159. 
> I would personally use the Chao estimator because it is more sensitive to the presence of rare species. 


### Rarefaction

In the R code chunk below, please do the following:

1. Calculate observed richness for all samples in `soilbac`,

2. Determine the size of the smallest sample,

3. Use the `rarefy()` function to rarefy each sample to this level,

4. Plot the rarefaction results, and

5. Add the 1:1 line and label.


```{r}
soilbac.S <- S.obs(soilbac.t)
min.N <- min(rowSums(soilbac.t))
S.rarefy <- rarefy(x = soilbac.t, sample = min.N, se = TRUE)
rarecurve(x = soilbac.t, step = 20, col = "blue", cex = 0.6, las = 1)
abline(0, 1, col = 'red')
text(1500, 1500, "1:1", pos = 2, col = 'red')


```


## 4) SPECIES EVNENNESS
Here, we consider how abundance varies among species, that is, **species evenness**.

### Visualizing evenness: the rank abundance curve (RAC)
One of the most common ways to visualize evenness is in a **rank-abundance curve** (sometime referred to as a rank-abundance distribution or Whittaker plot).
An RAC can be constructed by ranking species from the most abundant to the least abundant without respect to species labels (and hence no worries about 'ties' in abundance). 

In the R code chunk below, do the following:

1. Write a function to construct a RAC,

2. Be sure your function removes species that have zero abundances, 

3. Order the vector (RAC) from greatest (most abundant) to least (least abundant), and 

4. Return the ranked vector

```{r}

RAC <- function(x = ""){
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  as.data.frame(lapply(x.ab.ranked, unlist))
  return(x.ab.ranked)
}


```

Now, let us examine the RAC for `site1` of the BCI data set.

In the R code chunk below, do the following:

1. Create a sequence of ranks and plot the RAC with natural-log-transformed abundances,

2. Label the x-axis "Rank in abundance" and the y-axis "log(abundance)"

```{r}

plot.new()
site1 <- BCI[1, ]

rac <- RAC(x = site1)
ranks <- as.vector(seq(1, length(rac)))
opar <- par(no.readonly = TRUE)
par(mar = c(5.1, 5.1, 4.1, 2.1))
plot(ranks, log(rac), type = 'p', axes = F,
     xlab = "Rank in abundance", ylab = "Abundance",
     las = 1, cex.lab = 1.4, cex.axis = 1.25)
box()
axis(side = 1, labels = T, cex.axis = 1.25)
axis(side = 2, las = 1, cex.axis = 1.25,
     labels = c(1, 2, 5, 10, 20), at = log(c(1, 2, 5, 10, 20)))




```

***Question 5***: What effect does visualizing species abundance data on a log-scaled axis have on how we interpret evenness in the RAC?

> ***Answer 5***: Viewing the data on a log-scale adjusts the spread of the data to allow for clearer visualization of rare species without the most abundant species over crowding the graph and making it difficult to assess eveness.  


Now that we have visualized unevennes, it is time to quantify it using Simpson's evenness ($E_{1/D}$) and Smith and Wilson's evenness index ($E_{var}$).

### Simpson's evenness ($E_{1/D}$)

In the R code chunk below, do the following:

1. Write the function to calculate $E_{1/D}$, and

2. Calculate $E_{1/D}$ for `site1`.

```{r}

SimpE <- function(x = ""){
  S <- S.obs(x)
  x = as.data.frame(x)
  D <- diversity(x, "inv")
  E <- (D)/S
  return(E)
}

site1 <- BCI[1, ]
SimpE(site1)



```

### Smith and Wilson's evenness index ($E_{var}$)

In the R code chunk below, please do the following:

1. Write the function to calculate $E_{var}$,

2. Calculate $E_{var}$ for `site1`, and

3. Compare $E_{1/D}$ and $E_{var}$.

```{r}

Evar <- function(x){
  x <- as.vector(x[x > 0])
  1 - (2/pi) * atan(var(log(x)))
}

Evar(site1)
SimpE(site1)

```

***Question 6***: Compare estimates of evenness for `site1` of BCI using $E_{1/D}$ and $E_{var}$.
Do they agree? If so, why? If not, why? What can you infer from the results.

> ***Answer 6***: They do not agree. Simpson's evenness was roughly 0.42 and Smith and Wilson's evenness was roughly 0.51. They do not agree because Simpson's evenness is likely being biased by the most abundant species whereas the S&W index adjusts their calculations to control for the impact of very abundant species. This means that there are likely a few dominant species in the community that are skewing the simpson's index. 


## 5) INTEGRATING RICHNESS AND EVENNESS: DIVERSITY METRICS

So far, we have introduced two primary aspects of diversity, i.e., richness and evenness.
Here, we will use popular indices to estimate diversity, which explicitly incorporate richness and evenness.
We will write our own diversity functions and compare them against the functions in `vegan`. 

### Shannon's diversity (a.k.a., Shannon's entropy)

In the R code chunk below, please do the following:

1. Provide the code for calculating H' (Shannon's diversity),

2. Compare this estimate with the output of `vegan`'s diversity function using method = "shannon".

```{r}

ShanH <- function(x = ""){
  H = 0
  for (n_i in x){
    if(n_i > 0) {
      p = n_i / sum(x)
       H = H - p*log(p)
  }
}
return(H)
}

ShanH(site1)

diversity(site1, index = "shannon")

```

### Simpson's diversity (or dominance)

In the R code chunk below, please do the following:

1. Provide the code for calculating D (Simpson's diversity),

2. Calculate both the inverse (1/D) and 1 - D,

2. Compare this estimate with the output of `vegan's` diversity function using method = "simp".

```{r}

SimpD <- function(x = ""){
  D = 0
  N = sum(x)
  for (n_i in x){
    D = D + (n_i^2)/(N^2)
  }
  return(D)
}


D.inv <- 1/SimpD(site1)
D.sub <- 1-SimpD(site1)

diversity(site1, "inv")
diversity(site1, "simp")




```

### Fisher's $\boldsymbol\alpha$

In the R code chunk below, please do the following:

1. Provide the code for calculating Fisher's $\boldsymbol\alpha$,

2. Calculate Fisher's $\boldsymbol\alpha$ for `site1` of BCI.

```{r}

rac <- as.vector(site1[site1 > 0])
invD <- diversity(rac, "inv")
invD

Fisher <- fisher.alpha(rac)
Fisher


```

***Question 7***: 
How is Fisher's $\boldsymbol\alpha$ different from $E_{H'}$ and $E_{var}$? 
What does Fisher's $\boldsymbol\alpha$ take into account that $E_{H'}$ and $E_{var}$ do not?

> ***Answer 7***:Fisher's alpha assumes a log-series distribution of abundances of species which is common in diverse communities. This allows it to focus less on species evenness and more on species richness which can help control for biases caused by having a few highly abundant species in a community. This bias is still present in Shannon's H and Simpson's D although less so in the former than the latter. 

## 6) HILL NUMBERS 

Remember that we have learned about the advantages of Hill Numbers to measure and compare diversity among samples. 
We also learned to explore the effects of rare species in a community by examining diversity for a series of exponents $q$. 

***Question 8***: Using `site1` of BCI and `vegan` package, a) calculate Hill numbers for $q$ exponent 0, 1 and 2 (richness, exponential Shannon's entropy, and inverse Simpson's diversity). b) Interpret the effect of rare species in your community based on the response of diversity to increasing exponent $q$. 
```{r}
site1_rel <- site1 / sum(site1)  # Normalize manually

hill_number <- function(abundance, q) {
  if (q == 0) {
    return(sum(abundance > 0))  # Species richness (q = 0)
  } else if (q == 1) {
    return(exp(-sum(abundance * log(abundance), na.rm = TRUE)))  # Shannon diversity (q = 1)
  } else {
    return((sum(abundance^q))^(1 / (1 - q)))  # Simpson diversity (q = 2)
  }
}

hill_q0 <- hill_number(site1_rel, q = 0)  # Species Richness
hill_q1 <- hill_number(site1_rel, q = 1)  # Shannon Diversity
hill_q2 <- hill_number(site1_rel, q = 2)  # Simpson Diversity

# Combine into a named vector
hill_numbers <- c(q0 = hill_q0, q1 = hill_q1, q2 = hill_q2)

# Print results
print(hill_numbers)





```
> ***Answer 8a***: q0 = 93.00000, q1 = 55.61270, q2 = 39.41555
>
> ***Answer 8b***: Based on the hill numbers it appears that there is a large number of rare species in this community with a small number of dominant species. This is evident becuase both q1 and q2 are significantly smaller than q0 and q1 and q2 are more strongly influenced by evenness whereas q0 is based solely on species abundance. 
>

##7) MOVING BEYOND UNIVARIATE METRICS OF $\alpha$ DIVERSITY

The diversity metrics that we just learned about attempt to integrate richness and evenness into a single, univariate metric.
Although useful, information is invariably lost in this process. 
If we go back to the rank-abundance curve, we can retrieve additional information -- and in some cases -- make inferences about the processes influencing the structure of an ecological system.

## Species abundance models

The RAC is a simple data structure that is both a vector of abundances.
It is also a row in the site-by-species matrix (minus the zeros, i.e., absences). 

Predicting the form of the RAC is the first test that any biodiversity theory must pass and there are no less than 20 models that have attempted to explain the uneven form of the RAC across ecological systems.

In the R code chunk below, please do the following:

1. Use the `radfit()` function in the `vegan` package to fit the predictions of various species abundance models to the RAC of `site1` in BCI,

2. Display the results of the `radfit()` function, and 

3. Plot the results of the `radfit()` function using the code provided in the handout.

```{r}

RACresults <- radfit(site1)
RACresults
plot.new()
plot(RACresults, las = 1, cex.lab = 1.4, cex.axis = 1.25)

```


***Question 9***: Answer the following questions about the rank abundance curves:
a) Based on the output of `radfit()` and plotting above, discuss which model best fits our rank-abundance curve for `site1`?
b) Can we make any inferences about the forces, processes, and/or mechanisms influencing the structure of our system, e.g., an ecological community?

> ***Answer 9a***: Based on the rank abundance curves generated above, the Mandelbrot model appears to best fit the rank abundance curve for site1. 
>
> ***Answer 9b***: After further investigation into the Mandelbrot model, it suggests that site1 has a few dominant species with a large number of rare species. This indicates that competitive exclusion is likely playing a large role in shaping the structure of this community. 
>


***Question 10***: Answer the following questions about the preemption model:
a.  What does the preemption model assume about the relationship between total abundance (*N*) and total resources that can be preempted?
b.  Why does the niche preemption model look like a straight line in the RAD plot?

> ***Answer 10a***: The preemption model assumes that there is a fixed number of resources which then determines the abundance of species. Each species will secure a fixed quantity of the resource. The first species gets the largest share of the resources and then each subsequent species gets a smaller and smaller share. 
>
> ***Answer 10b***: Each species is allocated a slightly smaller share of the resources from species 1 until the last species in the ecosystem. This results in a linear relationship between resource quantity and species abundance because they are directly correlated. 
>


***Question 10***: Why is it important to account for the number of parameters a model uses when judging how well it explains a given set of data?

> ***Answer 11***: The more parameters a model has, the better it will fit the data set, but over-parametizing a model can also end up missrepresenting the dataset. Therefore, it is important to know how many parameters are included in a model so that it can be assessed for the likeliness that it is an accurate representation of the dataset. Too few parameters or too many parameters and this likeliness decreases. 
>

## SYNTHESIS

1.  As stated by Magurran (2004) the  ${D = } \sum p_i^2$ derivation of Simpson's Diversity only applies to communities of infinite size. 
For anything but an infinitely large community, Simpson's Diversity index is calculated as ${D = } \sum \frac{n_i(n_i -1)} {N(N-1)}$. 
Assuming a finite community, calculate Simpson's D, 1 - D, and Simpson's inverse (i.e. 1/D) for `site 1` of the BCI site-by-species matrix. 

```{r}

SimpD2 <- function(x = ""){
  D = 0
  N = sum(x)
  for (n_i in x){
    D = D + (n_i * (n_i - 1)) / (N * (N - 1))
  }
  return(D)
}

D2 <- SimpD2(site1)
D2.inv <- 1/SimpD2(site1)
D2.sub <- 1- SimpD2(site1)


```

2.  Along with the rank-abundance curve (RAC), another way to visualize the distribution of abundance among species is with a histogram (a.k.a., frequency distribution) that shows the frequency of different abundance classes.
For example, in a given sample, there may be 10 species represented by a single individual, 8 species with two individuals, 4 species with three individuals, and so on.
In fact, the rank-abundance curve and the frequency distribution are the two most common ways to visualize the species-abundance distribution (SAD) and to test species abundance models and biodiversity theories.
To address this homework question, use the R function **hist()** to plot the frequency distribution for `site 1` of the BCI site-by-species matrix, and describe the general pattern you see.

```{r}

# Create a histogram for species abundance distribution (evenness)
hist(rac, 
     breaks = 20,            # Adjust bin size for clarity
     col = "lightblue",      # Color of bars
     border = "black",       # Border color
     main = "Species Evenness Histogram for Site1", 
     xlab = "Species Abundance", 
     ylab = "Frequency",
     cex.axis = 1.2,         # Increase axis label size
     cex.lab = 1.3,          # Increase axis title size
     cex.main = 1.5)         # Increase main title size

```
> The general pattern that I see in this histogram is that there is a very small number of species that were observed many times and a large number of species that were observed very few times. 

3.  We asked you to find a biodiversity dataset with your partner.
This data could be one of your own or it could be something that you obtained from the literature. 
Load that dataset.
How many sites are there?
How many species are there in the entire site-by-species matrix?
Any other interesting observations based on what you learned this week?

```{r}

library(dplyr)
library(tidyr)
library(tibble)

tree <-read.csv("https://raw.githubusercontent.com/anna-l-2/QB_biodiversity_project_EH/d5b1465aaa135077652b5f5f69fd568d4d47b496/TREE.csv")
print(tree)

site_species_df <- tree %>%
  count(PLOT, SPCD) %>%  # Count species occurrences per plot
  pivot_wider(names_from = SPCD, values_from = n, values_fill = list(n = 0))  # Convert to matrix


#necessary functions 
#observed richness
S.obs <- function(x = ""){
  rowSums(x > 0) *1 
} 
#Good's coverage
GC <- function (x = ""){1- (rowSums(x ==1)/rowSums(x))}
#chao1
chao1 <- function(x ="") {
  S.obs(x) + (sum(x==1)^2)/(2*sum(x ==2))
}
#chao2
chao2 <- function(site = "", SbyS = "") {
  SbyS = as.data.frame(SbyS)
  x = SbyS[site,]
  SbyS.pa <- (SbyS >0) *1
  Q1 = sum(colSums(SbyS.pa) ==1)
  Q2 = sum(colSums(SbyS.pa) ==2)
  chao2 = S.obs(x) +(Q1^2)/(2*Q2)
  return(chao2)
}
#Shannon's H
shanH <-function (x = ""){
  H =0
  for (n_i in x){
    if (n_i > 0) {
      p = n_i/sum(x)
      H = H-p*log(p)
    }
  }
  return(H)
}

#Evar
Evar <- function(x){
  x <-as.vector(x[x >0])
  1-(2/pi)*atan(var(log(x)))
}
#ACE
ace <- function (x = "", thresh =10){
  x <- x[x >0]
  S.abund <-length(which (x > thresh))
  S.rare <- length(which (x <= thresh))
  singlt <- length(which (x ==1))
  N.rare <-sum(x[which (x <= thresh)])
  C.ace <-1-(singlt/N.rare)
  i <- c(1:thresh)
  count <- function (i, y){
    length(y[y == i])
  }
  a.1 <- sapply(i, count, x)
  f.1 <-(i * (i-1))*a.1
  G.ace <- (S.rare/C.ace)*(sum(f.1)/(N.rare*(N.rare-1)))
  S.ace <- S.abund +(S.rare/C.ace) +(singlt/C.ace) *max(G.ace,0)
  return(S.ace)
}

#RAC
RAC <-function (x = "") {
  x.ab = x[x>0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  as.data.frame(lapply(x.ab.ranked, unlist))
  return(x.ab.ranked)
}

#SimpD
SimpD = function(x = ""){
  D=0
  N=sum(x)
  for (n_i in x){
    D =D +(n_i^2)/(N^2)
  }
  return(D)
}

#simpE
simpE <- function (x = ""){
  S <-S.obs(x)
  x = as.data.frame(x)
  D <- diversity(x,"inv")
  E = (D)/S
}

#Overview
length(unique(tree$SPCD)) #109 species in the Tree dataset
length(unique(tree$PLOT)) #6586 plots in the Tree dataset

plot1 <- site_species_df[1, 2:110]


#Observed Richness
S.obs(plot1) #Observed richness = 2
obsRich.all <- S.obs(site_species_df) #Observed richness ranges from [2,12] for the entire dataset
range(obsRich.all)

#Good's Coverage
GC(plot1) #.889 GC
GC.tree.all <-GC(site_species_df) #mean = .6391--Anna
mean(GC.tree.all) #Updated mean=0.998? --Elaine

chao2.tree.all <- chao1(site_species_df) #range [17048.85, 17058.85 mean 17051.29
range(chao2.tree.all)
mean(chao2.tree.all)
print(chao2.tree.all)




```

## SUBMITTING YOUR ASSIGNMENT
Use Knitr to create a PDF of your completed 5.AlphaDiversity_Worksheet.Rmd document, push it to GitHub, and create a pull request.
Please make sure your updated repo include both the pdf and RMarkdown files.

Unless otherwise noted, this assignment is due on **Wednesday, January 29^th^, 2025 at 12:00 PM (noon)**.