---
title: '6\. Worksheet: Among Site (Beta) Diversity -- Part 2'
author: "Elaine Hoffman; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we continue to explore concepts, statistics, and visualizations related to $\beta$-diversity.
Now that you know how to formally quantify $\beta$-diversity, we will learn how to test hypotheses about $\beta$-diversity using multivariate statistics. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, you should **push** this file to your GitHub repo, at whatever stage you are. This will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Posit.cloud workspace: `/cloud/project/QB-2025/Week4-Beta/`
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file (**6.BetaDiversity_2_Worksheet.Rmd**) 
with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**6.BetaDiversity_2_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 12^th^, 2025 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your `Week4-Beta/` folder. 
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list = ls())
setwd("/cloud/project/Week4-Beta")
library(vegan)
library(ade4)
library(indicspecies)


```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, load the `doubs` dataset from the `ade4` package

```{r}
# note, pleae do not print the dataset when submitting
data(doubs)


```
## 3) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}
fish <- doubs$fish[-8, ]

#Create factors vector
quality <- c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))

#Run PERMANOVA with adonis2 function
adonis2(fish ~ quality, method = "bray", permutations = 999)

indval <- multipatt(fish, cluster = quality, func = "IndVal.g", control = how(nperm = 999))
summary(indval)

fish.rel <- decostand(fish, method = "total")
phi <- multipatt(fish.rel, cluster = quality, func = "r.g",
                 control = how(nperm = 999))

summary(phi)




```

***Question 1***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?
Are the different analyses consistent with one another and do they agree with the visualizations (heat maps, cluster dendograms, ordinations) that you created?

> ***Answer 1***: These analyses revealed that there is a significant relationship between habitat quality andd fish species composition. The PERMANOVA found that water quality predicts roughly 46% of the variation in fish composition along the water quality gradient in the river between the three groups (L, M, H). The IndVal and phi coefficient supported this conclusion by identifying species that grouped within each category. The results were very similar, but the phi coefficient provided a greater resolution to the groupings by placing more species in single categories rather than spanning 2 as seen for some species in the IndVal analysis. This aligns with the visualizations produced last week that indicated that water quality was likely significantly related to fish species distributions. 


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
#Define matrices
fish.dist <- vegdist(doubs$fish[-8, ], method = "bray")
env.dist <- vegdist(scale(doubs$env[-8, ]), method = "euclid")

#Mantel test
mantel(fish.dist, env.dist)

```

***Question 2***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? 
How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 2***:  The results of the Mantel test indicate that there is a moderate to strong relationship (r = 0.604) between fish diversity and stream environmental conditions and was strongly statistically signficant (0.01). This supports my hypothesis that stream quality is influencing fish communities.  

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   
```{r}

#calculate bray-curtis
fish.db <- vegdist(fish, method = "bray")

#Define environmental matrix
env.chem <- as.matrix(doubs$env[-8, 5:11])

#Perform dbRDA
doubs.dbrda <- dbrda(fish.db ~ ., as.data.frame(env.chem))
ordiplot(doubs.dbrda)

#First we will model only the intercept
doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))

#Note there are no vectors here (we didn't constrain anything)
#Therefore, the axes suggest this is a simple MDS (i.e. PCoA)
ordiplot(doubs.dbrda.mod0)

#Next, we will model the full model, with all explanatory variables
doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))

#Now we will step through all combinations of explanatory variables in our model
#The function returns the model with the lowest AIC value
doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)

#Lets look at the model that was selected
doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)

#Permutation tests to evaluate significance 
permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[,c(4,6,7)], perm = 999)

#Calculate Explained Variation
dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1] / 
                             sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100

dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] / 
                             sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100

#Define plot parameters
par(mar = c(5, 5, 4, 4) + 0.1)

#Initiate plot
plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1),
     ylim = c(-1.1, 2.7), xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)", 
                                       sep = ""), ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
     cex.axis = 1.2, axes = FALSE)


# Add axes
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

#Add points and labels 
points(scores(doubs.dbrda, display = "wa"),
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"),
     labels = row.names(scores(doubs.dbrda, display = "wa")))

#Add environmental vectors 
vectors <- scores(doubs.dbrda, display = "bp")

#row.names(vectors) <- rownames(vectors)
arrows(0, 0, vectors[,1], vectors[, 2],
       lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[, 2], pos = 3, 
     labels = row.names(vectors))
axis(side = 3, lwd.ticks = 2, cex.axis =1.2, las = 1, col = "red", lwd = 2.2, 
     at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2, 
     at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))



```

***Question 3***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 3***: The environmental variables that seem to be contributing to variation in fish community structure are nitrogen, oxygen, and bdo. 

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}

#Remember, our environmental model uses oxy, bdo, and nit has R2 of 0.53
doubs.dbrda$anova

#Let's create a matrix model for our environmental data
env.mod <- model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]

#First, we will weight each site by its relative abundance
rs <- rowSums(fish)/sum(fish)

#Next, we will perform PCNM
doubs.pcnmw <- pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T)

#PCNM can return negative eigenvalues, but only the eigen vectors associated with the positive eigenvalue are meaningful
doubs.pcnmw$values > 0

doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

#Because this is another dbRDA, we could visualize the biplot
#showing how each vector explains variation across sites
plot(step.pcnm)

#The object 'step.pcnm' now contains the selected model.
step.pcnm$anova

#We can now construct a spatial model using only the selected PCNM axes.
space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 + 
                            PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1]

#First conduct constrained ordinations
doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)

#Next construct partial constrained ordinations
doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod))

#Next test for significance of the dbRDA fractions.
permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)

#Using the built-in varpart() function
doubs.varpart <- varpart(fish.db, env.mod, space.mod)
doubs.varpart

par(mar = c(2, 2, 2, 2))
plot(doubs.varpart)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation Partioning of\nDoubs Fish Diversity", side = 3, line = -3)




```

***Question 4***: Interpret the variation partitioning results.   

> ***Answer 4***: 24% of the variation is explained by space alone, 13% of variation is explained by environment alone, 37% of the variation is explained by space and environment together. 

## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the hypothesis-testing tools introduced in the beta diversity module. 
Interpret the findings of your data with respect to principles of biodiversity. 

```{r}
# Load necessary libraries
library(vegan)
library(tibble)
library(lattice)
library(readxl)
library(dplyr)
library(tidyr)


#Trees
tree <-read.csv("https://raw.githubusercontent.com/anna-l-2/QB_biodiversity_project_EH/d5b1465aaa135077652b5f5f69fd568d4d47b496/TREE.csv")
tree.species.df <-data.frame(Plot_ID = c(tree$PLOT),
                             SPCD =c(tree$SPCD))
#str(tree.species.df)
as.numeric()






#Myco
url <- "https://raw.githubusercontent.com/anna-l-2/QB_biodiversity_project_EH/main/MycoType_ref2.xlsx"
# Create a temporary file to store the download
temp_file <- tempfile(fileext = ".xlsx")
# Download the file
download.file(url, temp_file, mode = "wb")
# Read the Excel file
myco <- read_excel(temp_file)
myco<- myco[,-c(3:7)]


#invasive
url <- "https://raw.githubusercontent.com/anna-l-2/QB_biodiversity_project_EH/main/INVASIVE.xlsx"
# Create a temporary file to store the download
temp_file2 <- tempfile(fileext = ".xlsx")
# Download the file
download.file(url, temp_file2, mode = "wb")
# Read the Excel file
invasive <- read_excel(temp_file2)
invasive.df <-as.data.frame(invasive$PLOT) 



#References

#Plot ID code
url <- "https://raw.githubusercontent.com/anna-l-2/QB_biodiversity_project_EH/main/PLOT.xlsx"
# Create a temporary file to store the download
temp_file3 <- tempfile(fileext = ".xlsx")
# Download the file
download.file(url, temp_file3, mode = "wb")
# Read the Excel file
plotID.master <- read_excel(temp_file3)


#Species ID
# GitHub raw file URL
url <- "https://raw.githubusercontent.com/anna-l-2/QB_biodiversity_project_EH/main/REF_SPECIES.xlsx"
# Create a temporary file to store the download
temp_file4 <- tempfile(fileext = ".xlsx")
# Download the file
download.file(url, temp_file4, mode = "wb")
# Read the Excel file
speciesID.master <- read_excel(temp_file4)




untree.species <- unique(tree.species.df$SPCD)
unmyco.species <-unique(myco$SPCD)
tree.species.df <- tree.species.df %>%
  filter(!SPCD %in% c(999, 998)) 



#add myco type
tree.species.df$PlotStatus <- NA


# Loop through rows of tree.species.df
for (i in seq_len(nrow(tree.species.df))) {  
  if (tree.species.df$Plot_ID[i] %in% invasive$PLOT) {  
    tree.species.df$PlotStatus[i] <- "invasive"  
  } else {  
    tree.species.df$PlotStatus[i] <- "non-invasive"  
  }
}
#Site by Species
tree.species.df.1 <-data.frame(Plot_ID = c(tree$PLOT),
                               Species_ID =c(tree$SPCD))



tree.ss.df <- as.data.frame.matrix(table(tree.species.df.1$Plot_ID, tree.species.df.1$Species_ID))
tree.ss.df <- rownames_to_column(tree.ss.df, var = "Plot_ID")
#print(tree.ss.df)

tree.species.only.df <- tree.ss.df[, 2:ncol(tree.ss.df)]


#Create Factors vector

factor <- factor(tree.species.df$PlotStatus)

length(factor)
nrow(tree.species.only.df)

factor_trimmed <- factor[1:nrow(tree.species.only.df)]

#adonis2(tree.species.only.df ~ factor_trimmed, method = "bray", permutations = 999)





```
